Effects of PCSK9 Inhibition With Evolocumab on Lipoprotein Metabolism in subjects with elevated Lp(a)

### 2019-10-04 manuscript.Rmd
add discussion on apo(a) composition

Lipoprotein(a) [Lp(a)] is an independent risk factor for cardiovascular diseases, and often considered as a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail produced by *LPA*. 
Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration. 
Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].
PCSK9 binds to hepatic low density lipoprotein (LDL) receptors and mediates their degradation in lysosome.
Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018]. 
As shown in a recent meta-analysis, useage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
These observations suggested existence of non-LDL receptor mediated catabolism of Lp(a). 
In the current study, we aim to explore the alternative pathways beyond LDL receptor-mediated catabolism of Lp(a).
To this end, we studied the metabolic effects of PCSK9 inhibition with evolocumab in 30 subjects (placebo 14; evolocumab 16) with elevated Lp(a). 
We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
Our further analysis revealed that evolocumab treatment not only reduced LDL particle concentrations, but also changed the distribution between LDL particles with and without apo(a).

@@

In the current study, evolocumab treatment for 16 weeks resulted in on average 18.9% reduction in Lp(a) compared with placebo.
Our detailed lipoprotein profiling revealed that treatment of evolocumab substantially reduced concentrations of VLDL, IDL and LDL particles. 
Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
The next question we came up with is "why we observed large effect of evolocumab on VLDL, IDL and LDL particle concentrations but relatively small effect on Lp(a)?" 
To address the question, we analyzed the composition of LDL particles. 
Because Lp(a) is often thought to be comprised of a LDL particle and a covalently bound apo(a), we assumed that a pool of LDL particles contained LDL particles with and without apo(a) tail. 
At baseline, we identified that on average 40%-50% of large, medium and small LDL particles had apo(a) tail. 
After evolocumab treatment, we identified that the proportion of LDL with apo(a) in all three pools of LDL particles increased to about 70%-75%. 
The composition shifting to relatively more LDL particles with apo(a) indicated that the reduction of LDL particles may mainly correspond to the reduction of LDL particles without apo(a). 
Indeed Lp(a) needs to compete with LDL particles for binding to the LDL receptor, and is preferred for uptake by the LDL receptor when the LDL particle concentration is low [@Raal2016].  
We hypothesized that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
Besides LDL receptor, we acknowledge that VLDL receptor can also mediate Lp(a) catabolism [@Argraves1997]. 
In our analysis, we observed substantial reduction in VLDL particle concentrations after treatment of evolocumab. 
Since PCSK9 can induce the degradation of LDL receptor as well as VLDL receptor, evolocumab treatment must enhance activity of both receptors [@Poirier2008]. 
Interestingly, we observed that most of patients receiving evolocumab ended in very low concentration (below the detection limit) of extremely large and very large VLDL particles, suggesting enhancement of VLDL receptor activity.

>>>>

Lipoprotein(a) [Lp(a)] is an independent risk factor for cardiovascular diseases, and often considered as a low-density lipoprotein (LDL)-like particle with an apolipoprotein(a) [apo(a)] tail produced by *LPA*. 
Unlike LDL-cholesterol lowering, limited therapeutic options are available to reduce plasma Lp(a) concentration. 
Proprotein convertase subtilisin/kexin type 9 (PCSK9) monoclonal antibody such as evolocumab has been shown to lower Lp(a) [@Raal2014; @Watts2017; @Shapiro2019].
PCSK9 binds to hepatic low density lipoprotein (LDL) receptors and mediates their degradation in lysosome.
Evolocumab (also called AMG 145) is a fully human monoclonal antibody that targets PCSK9 and prevents it from binding to the LDL receptor. 
Although evolocumab-induced Lp(a) reduction was linked to enhanced LDL receptor mediated uptake and attenuated Lp(a) production, the mechanism of Lp(a) lowering by PCSK9 inhibition is unknown [@Romagnuolo2015; @Raal2016; @Watts2018]. 
As shown in a recent meta-analysis, usage of statins leads to increasing of Lp(a) concentration [@Tsimikas2019]. 
More interestingly, decreasing Lp(a) concentration was seen in homozygous familial hypercholesterolaemia (FH) patients after treatment of evolocumab [@Raal2015]. 
These observations suggested existence of non-LDL receptor mediated catabolism of Lp(a). 
In the current study, we aim to explore the alternative pathways beyond LDL receptor-mediated catabolism of Lp(a).
To this end, we studied the metabolic effects of PCSK9 inhibition with evolocumab in 30 subjects (placebo 14; evolocumab 16) with elevated Lp(a). 
We identified that despite evolocumab treatment for 16 weeks resulted in 18.9% reduction of Lp(a), coupled with substantial reduction of VLDL, IDL and LDL particles.   
Our further analysis revealed that evolocumab treatment not only reduced LDL particle concentrations, but also changed the distribution between LDL particles with and without apo(a).

@@

In the current study, evolocumab treatment for 16 weeks resulted in on average 18.9% reduction in Lp(a) compared with placebo.
Our detailed lipoprotein profiling revealed that treatment of evolocumab substantially reduced concentrations of VLDL, IDL and LDL particles. 
Our observation is consistent with the previous study, in which evolocumab treatment not only reduced concentrations of LDL particles, but also substantially reduced triglyceride-rich lipoprotein particles [@Toth2018].
The next question we came up with is "why we observed large effect of evolocumab on VLDL, IDL and LDL particle concentrations but relatively small effect on Lp(a)?" 
To address the question, we analyzed the composition of LDL particles. 
Because Lp(a) is often thought to be comprised of a LDL particle and a covalently bound apo(a), we assumed that a pool of LDL particles contained LDL particles with and without apo(a) tail. 
At baseline, we identified that on average about 50% of large, medium and small LDL particles had apo(a) tail. 
Evolocumab treatment not only decreased concentrations of LDL particles, but also resulted in 10-20% increasing in LDL particles with apo(a). 
The composition shifting to relatively more LDL particles with apo(a) indicated that LDL particles with apo(a) were less preferably cleared from the circulation compared to their counterparts without apo(a). 
Indeed Lp(a) needs to compete with LDL particles for binding to the LDL receptor, and is preferred for uptake by the LDL receptor when the LDL particle concentration is low [@Raal2016].  
We hypothesized that the LDL particles with and without apo(a) tail may have different affinity for binding to the LDL receptor. 
Besides LDL receptor, we acknowledge that VLDL receptor can also mediate Lp(a) catabolism [@Argraves1997]. 
In our analysis, we observed substantial reduction in VLDL particle concentrations after treatment of evolocumab. 
Since PCSK9 can induce the degradation of LDL receptor as well as VLDL receptor, evolocumab treatment must enhance activity of both receptors [@Poirier2008]. 
Interestingly, we observed that most of patients receiving evolocumab ended in very low concentration (below the detection limit) of extremely large and very large VLDL particles, suggesting enhancement of VLDL receptor activity.
